We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-19 paper/paper.tex
ref
references

\begin{abstract}
We use numerical simulations to build an explicit probability
distribution for the asphericity in the satellite distribution around
galaxies similar to the Local Group (LG) in the Lambda Cold Dark
Matter (LCDM) paradigm. 
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ among the simulations suggesting a strong
influence of the typical Dark Matter (DM) halo mass in the LG samples.
However, there are three common trends.
First, at most of $2\%$ the pairs are expected to have satellite
distributions with the same asphericity as the LG; second,
between $27\%$ to $56\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $3\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in the M31 galaxy have a rather typical
distribution in the LCDM context, while the MW distribution is
atypical in this respect.  
These results place the LG at the level of a $3\sigma$ outlier in the
LCDM paradigm. 
Understand this atipicality will require quantifying the probability
distribution for the satellite distribuition asphericity as a function
of halo mass and large scale environment. 
The approach presented can facilitate this study and other comparisons
between different numerical setups and choices to study satellites
around LG pairs in simulations.  
\end{abstract}

@@

The spatial distribution of satellite galaxies around our Milky Way
(MW) and the M31 galaxy provides a stringent test for structure
formation theories in a explicit cosmological context
\citep{2015ApJ...815...19P, 2016MNRAS.457.1931S, 2016arXiv160601516L}.  
The presence of a Vast Polar Structure of satellites around the Milky
Way has been stablished in the last decade \citep{2013MNRAS.435.1928P}.
This structure is present in the $11$ brightest "classical" dwarf
galaxies and has been confirmed as fainter satellites have been
detected \citep{2014MNRAS.441.2124B}. %XXXXXXX  
In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
%homogeneous parent sample
%comentario: yo no describiría la poblacion como homogenea porque la
%mayoría están entre M31 y la MW la poblción está "loopsided"
total poulation of $27$ satellites.\citep{2013ApJ...766..120C,
2013Natur.493...62I} 
However, taking into account only the $11$ brightest satellites (as a
comparison with the MW) there is no evidence of any planar structure
in M31.

@@

>>>>

\begin{abstract}
We use numerical simulations to build an explicit probability
distribution for the asphericity in the satellite distribution around
galaxies similar to the Local Group (LG) in the Lambda Cold Dark
Matter (LCDM) paradigm. 
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ among the simulations suggesting a strong
influence of the typical Dark Matter (DM) halo mass in the LG samples.
However, there are three common trends.
First, at most of $2\%$ the pairs are expected to have satellite
distributions with the same asphericity as the LG; second,
between $27\%$ to $56\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $3\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in the M31 galaxy have a rather typical
distribution in the LCDM context, while the MW distribution is
atypical in this respect.  
These results place the LG at the level of a $3\sigma$ outlier in the
LCDM paradigm. 
We suggest that to understand this atipicality will require quantifying
the probability distribution for the satellite distribuition
asphericity as a function of halo mass and large scale environment. 
The approach presented here can facilitate that study and other comparisons
between different numerical setups and choices to study satellites
around LG pairs in simulations.  
\end{abstract}

@@

The spatial distribution of satellite galaxies around our Milky Way
(MW) and the M31 galaxy provides a stringent test for structure
formation theories in a explicit cosmological context
\citep{2015ApJ...815...19P, 2016MNRAS.457.1931S}
The presence of a Vast Polar Structure of satellites around the Milky
Way, first suggested by \cite{1976RGOB..182..241K} and
\cite{1976MNRAS.174..695L} as a Magellanic Plane, has been
stablished in the last decade \citep{2005A&A...431..517K,2013MNRAS.435.1928P}. 
This structure is present in the $11$ brightest "classical" dwarf
galaxies and has been confirmed as fainter satellites have been
detected \citep{2014MNRAS.441.2124B}. %XXXXXXX  
In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
total poulation of $27$ satellites.\citep{2013ApJ...766..120C,
2013Natur.493...62I}

@@

\section*{Acknowledgements} 
We acknowledge Universidad de los Andes and COLCIENCIAS (Project -
XXX) for the computing resources used in this project.
